### 2020-09-09 ART_Czarnik/Czarnik-PU.tex
Final version before printing and publishing online

\enlargethispage{.5\baselineskip}
\section[Problems in measuring the accuracy of stereotypes]{Problems in measuring the accuracy of stereotypes\footnote{It should be stressed that what I am interested in here is the extent to which stereotypical judgments as such are correct. It is a related but separate issue whether the process by which people arrive at their judgments is correct. For example, someone may arrive at a correct estimate of honesty in a given group of people, even though he overestimates honesty of people in general while underestimating honesty of that particular group relative to general public. See
%\label{ref:RNDq1DptTqYi9}(Jussim et al., 2016)
\parencite[][]{jussim_stereotype_2016} %
 for a discussion of processual models of judgment by Cronbach 
%\label{ref:RNDqEAFrZBRPd}(1955)
\parencite*[][]{cronbach_processes_1955} %
 and Kenny 
%\label{ref:RNDlQ5sgtU4cA}(Kenny, 1994).
\parencite*[][]{kenny_interpersonal_1994}. %
 }}
Assessing stereotype (in)accuracy involves three obvious steps
%\label{ref:RNDr97wsTQtKY}(Jussim, Crawford and Rubinstein, 2015):
\parencite[][]{jussim_stereotype_2015}:%

>>>>

\enlargethispage{.5\baselineskip}
\section[Problems in measuring the accuracy of stereotypes]{Problems in measuring the accuracy of stereotypes\footnote{\ It should be stressed that what I am interested in here is the extent to which stereotypical judgments as such are correct. It is a related but separate issue whether the process by which people arrive at their judgments is correct. For example, someone may arrive at a correct estimate of honesty in a given group of people, even though he overestimates honesty of people in general while underestimating honesty of that particular group relative to general public. See
%\label{ref:RNDq1DptTqYi9}(Jussim et al., 2016)
\parencite[][]{jussim_stereotype_2016} %
 for a discussion of processual models of judgment by Cronbach 
%\label{ref:RNDqEAFrZBRPd}(1955)
\parencite*[][]{cronbach_processes_1955} %
 and Kenny 
%\label{ref:RNDlQ5sgtU4cA}(Kenny, 1994).
\parencite*[][]{kenny_interpersonal_1994}. %
 }}
Assessing stereotype (in)accuracy involves three obvious steps
%\label{ref:RNDr97wsTQtKY}(Jussim, Crawford and Rubinstein, 2015):
\parencite[][]{jussim_stereotype_2015}:%

### 2020-09-09 ART_Dada_Piech/Dada-PU.tex
Final version before printing and publishing online

\begin{artengenv}{Jerzy Dadaczyński, Robert Piechowicz}
	{Maxime Bôcher's concept of complementary philosophy of mathematics}
	{Maxime Bôcher's concept of complementary philosophy of mathematics}
	{Maxime Bôcher's concept of complementary philosophy\\of mathematics}
	{Pontifical University of John Paul II in Krakow}
	{The main purpose of the present paper is to demonstrate that as early as 1904 pre-eminent American mathematician Maxime Bôcher was an adherent to the presently relevant argument of reasonableness, or even necessity of parallel development of two philosophical methods of reflection on mathematics, so that its essence could be more fully comprehended. The goal of the research gives rise to the question: what two types of philosophical deliberation on mathematics were proposed by Bôcher?}
	{philosophy, mathematics, Maxim Bôcher, metamathematics.}

>>>>

\begin{artengenv}{Jerzy Dadaczyński, Robert Piechowicz}
	{Maxime Bôcher's concept of complementary philosophy of mathematics}
	{Maxime Bôcher's concept of complementary philosophy of mathematics}
	{Maxime Bôcher's concept of complementary philosophy\\of mathematics}
	{Pontifical University of John Paul II in Krakow, Poland}
	{The main purpose of the present paper is to demonstrate that as early as 1904 pre-eminent American mathematician Maxime Bôcher was an adherent to the presently relevant argument of reasonableness, or even necessity of parallel development of two philosophical methods of reflection on mathematics, so that its essence could be more fully comprehended. The goal of the research gives rise to the question: what two types of philosophical deliberation on mathematics were proposed by Bôcher?}
	{philosophy, mathematics, Maxim Bôcher, metamathematics.}

### 2020-09-09 ART_Kawalec/Kawalec-PU.tex
A. Figarski correctons added. Sending to the authors
Final version before printing and publishing online

\section{Particularist account of the microRNAs discovery}
In 1993 two research teams led by Victor Ambros and Gary Ruvkun, respectively, published papers reporting a new kind of RNA particle: \textit{lin-4}. The well-known central dogma identified the main---and presumably exclusive---role of RNA as the template for protein sequences.\footnote{There is some discrepancy between the original formulation of the central dogma by F. Crick
%\label{ref:RNDTAwGDm5AsM}(1958, 1970)
\parencite*[][]{crick_central_1970} %
 and its ``lucid presentation'' 
%\label{ref:RNDUB41shCuDU}(Watson, 1965, pp.297–298)
\parencite[][pp.297–298]{watson_molecular_1965} %
 by S. Spiegelman and M. Hayashi 
%\label{ref:RNDVMNFyp8qd8}(1963).
\parencite*[][]{spiegelman_present_1963}. %
 While Crick's version, essentially, only precludes the information transfer from proteins to DNA or RNA, Spiegelman and Hayashi require one-directional information transfer from RNA to protein, excluding thereby two other routes that are possible in Crick's version. I am grateful to one of the Reviewers for pointing that out.} However, the RNA discovered in 1993 had a different function: it regulated gene expression without encoding any protein. What the two teams demonstrated was that in nematode \textit{Caenorhabditis elegans} (\textit{C. elegans} in short) the gene \textit{lin-4} negatively regulates translation of another gene \textit{lin-14} that, in turn, codes LIN-14 protein. When this mechanism is corrupted by a mutation of \textit{lin-4} causing its loss-of-function, \textit{C. elegans} reiterates early fates at late developmental stages, being unable to develop adult structures and to lay eggs 
%\label{ref:RNDl5HcelhTv8}(Lee, Feinbaum and Ambros, 1993).
\parencite[][]{lee_c_1993}. %
 The authors succinctly summarize two lines of evidence supporting the claim that \textit{lin-4} does not encode protein:

@@

Before my discussion of the sequence of discoveries related to microRNAs and the corresponding de Solla Price's growth curve, let me examine the origins and presumptions of the particularist approach to the advancement of science.
\begin{figure}[h!]
	\includegraphics[width=1\textwidth]{ART_Kawalec/Kawalec-img001bw.png}
	\caption{The annual growth of publication number on microRNAs. Source: own analysis of MEDLINE dataset for ``microRNAs'' as MeSH keyword.}\label{fig1kawalec}
\end{figure}

@@

\\
 \includegraphics[width=.5\textwidth]{ART_Kawalec/Kawalec-img002bw.jpg}  &
 \includegraphics[width=.5\textwidth]{ART_Kawalec/Kawalec-img003bw.png} \\
\multicolumn{2}{c}{{\centering\bfseries giant component}
}\\
\centering 6\% (of 436 authors) &
\centering\arraybackslash 70\% (of 38 703 authors)\\
\end{tabular}
\end{flushleft}
\caption{Dynamics of microRNA routine – the expansion of size and the giant component of collaborating authors. Source: own analysis of MEDLINE dataset for “microRNAs” as keyword.}\label{fig2kawalec}
\end{figure}

@@

Apparently, another important difference between \textit{lin-4} and \textit{let-7} that was noted early on was the recognition of the role of \textit{let-7} in oncogenesis and tumor suppression. Hence, \textit{let-7} has become an object of intense studies as it ``was considered as the most important miRNA for the cancer incidence and progression''
%\label{ref:RNDLaW8XsZykT}(Wang, Jiang and Xu, 2018).
\parencite[][]{wang_comprehensive_2018}. %
 The role of homologs of \textit{lin-4} in oncogenesis was noted later and was less apparent 
%\label{ref:RNDymkacnfBGk}(Sonoki et al., 2005).
\parencite[][]{sonoki_insertion_2005}. %
 The differences between \textit{lin-4} and \textit{let-7}, at least partly, explain the difference in their respective research dynamics as presented in figure \ref{fig3kawalec} which depicts the dynamics of topical expansion of both research lines against the annual growth of the number of publications.
\begin{figure}[h!]
	\includegraphics[width=1\textwidth]{ART_Kawalec/Kawalec-img004bw.png}
	\caption{Newcomer dynamics of MeSH keywords relative to the annual number of publications for the first two microRNAs: \textit{lin-4} and \textit{let-7}. Source: own analysis of MEDLINE datasets for ``\textit{lin-4}'' and ``\textit{let-7}'' as keywords, respectively. The number of hitherto unrecorded MeSH keywords (``newcomers'') for a given year was measured against the annual publication record for ``\textit{lin-4}'' and ``\textit{let-7}'', respectively. The trendlines were estimated as linear functions.}\label{fig3kawalec}
\end{figure}

>>>>

\section{Particularist account of the microRNAs discovery}
In 1993 two research teams led by Victor Ambros and Gary Ruvkun, respectively, published papers reporting a new kind of RNA particle: \textit{lin-4}. The well-known central dogma identified the main---and presumably exclusive---role of RNA as the template for protein sequences.\footnote{There is some discrepancy between the original formulation of the central dogma by F. Crick
%\label{ref:RNDTAwGDm5AsM}(1958, 1970)
\parencite*[][]{crick_central_1970} %
 and its ``lucid presentation'' 
%\label{ref:RNDUB41shCuDU}(Watson, 1965, pp.297–298)
\parencite[][pp.297–298]{watson_molecular_1965} %
 by S.~Spiegelman and M.~Hayashi 
%\label{ref:RNDVMNFyp8qd8}(1963).
\parencite*[][]{spiegelman_present_1963}. %
 While Crick's version, essentially, only precludes the information transfer from proteins to DNA or RNA, Spiegelman and Hayashi require one-directional information transfer from RNA to protein, excluding thereby two other routes that are possible in Crick's version. I am grateful to one of the Reviewers for pointing that out.} However, the RNA discovered in 1993 had a different function: it regulated gene expression without encoding any protein. What the two teams demonstrated was that in nematode \textit{Caenorhabditis elegans} (\textit{C. elegans} in short) the gene \textit{lin-4} negatively regulates translation of another gene \textit{lin-14} that, in turn, codes LIN-14 protein. When this mechanism is corrupted by a mutation of \textit{lin-4} causing its loss-of-function, \textit{C. elegans} reiterates early fates at late developmental stages, being unable to develop adult structures and to lay eggs 
%\label{ref:RNDl5HcelhTv8}(Lee, Feinbaum and Ambros, 1993).
\parencite[][]{lee_c_1993}. %
 The authors succinctly summarize two lines of evidence supporting the claim that \textit{lin-4} does not encode protein:

@@

Before my discussion of the sequence of discoveries related to microRNAs and the corresponding de Solla Price's growth curve, let me examine the origins and presumptions of the particularist approach to the advancement of science.
\begin{figure}[h!]
%	\includegraphics[width=1\textwidth]{ART_Kawalec/Kawalec-img001bw.png}
	\includegraphics[width=1\textwidth]{ART_Kawalec/Kawalec-img001.png}
	\caption{The annual growth of publication number on microRNAs. Source: own analysis of MEDLINE dataset for ``microRNAs'' as MeSH keyword.}\label{fig1kawalec}
\end{figure}

@@

\\
% \includegraphics[width=.5\textwidth]{ART_Kawalec/Kawalec-img002bw.jpg}  &
% \includegraphics[width=.5\textwidth]{ART_Kawalec/Kawalec-img003bw.png} \\
	\includegraphics[width=.5\textwidth]{ART_Kawalec/Kawalec-img002.jpg}  &
 	\includegraphics[width=.5\textwidth]{ART_Kawalec/Kawalec-img003.png} \\
\multicolumn{2}{c}{{\centering\bfseries giant component}
}\\
\centering 6\% (of 436 authors) &
\centering\arraybackslash 70\% (of 38 703 authors)\\
\end{tabular}
\end{flushleft}
\caption{Dynamics of microRNA routine – the expansion of size and the giant component of collaborating authors. Source: own analysis of MEDLINE dataset for “microRNAs” as keyword.}\label{fig2kawalec}
\end{figure}

@@

Apparently, another important difference between \textit{lin-4} and \textit{let-7} that was noted early on was the recognition of the role of \textit{let-7} in oncogenesis and tumor suppression. Hence, \textit{let-7} has become an object of intense studies as it ``was considered as the most important miRNA for the cancer incidence and progression''
%\label{ref:RNDLaW8XsZykT}(Wang, Jiang and Xu, 2018).
\parencite[][]{wang_comprehensive_2018}. %
 The role of homologs of \textit{lin-4} in oncogenesis was noted later and was less apparent 
%\label{ref:RNDymkacnfBGk}(Sonoki et al., 2005).
\parencite[][]{sonoki_insertion_2005}. %
 The differences between \textit{lin-4} and \textit{let-7}, at least partly, explain the difference in their respective research dynamics as presented in figure \ref{fig3kawalec} which depicts the dynamics of topical expansion of both research lines against the annual growth of the number of publications.
\begin{figure}[h!]
%	\includegraphics[width=1\textwidth]{ART_Kawalec/Kawalec-img004bw.png}
	\includegraphics[width=1\textwidth]{ART_Kawalec/Kawalec-img004.png}
	\caption{Newcomer dynamics of MeSH keywords relative to the annual number of publications for the first two microRNAs: \textit{lin-4} and \textit{let-7}. Source: own analysis of MEDLINE datasets for ``\textit{lin-4}'' and ``\textit{let-7}'' as keywords, respectively. The number of hitherto unrecorded MeSH keywords (``newcomers'') for a given year was measured against the annual publication record for ``\textit{lin-4}'' and ``\textit{let-7}'', respectively. The trendlines were estimated as linear functions.}\label{fig3kawalec}
\end{figure}

### 2020-09-09 ART_Liana/Liana-PU.tex
Final version before printing and publishing online

\begin{artplenv}{Zbigniew Liana}
	{Józefa Życińskiego koncepcja racjonalizmu umiarkowanego: epistemologiczna i~doksalogiczna funkcja podmiotowego \textit{commitment}}
	{Józefa Życińskiego koncepcja racjonalizmu umiarkowanego\ldots}
	{Józefa Życińskiego koncepcja racjonalizmu umiarkowanego: epistemologiczna i~doksalogiczna funkcja podmiotowego \tocname{commitment}}
	{Uniwersytet Papieski Jana Pawła II w~Krakowie\\	
	Copernicus Center for Interdisciplinary Studies\label{liana-anfang}}
	{Joseph Życiński's understanding of temperate rationalism: epistemological and doxalogical role of subjective commitment}
	{One of the main problems of modern rationalistic theories of science is the non-eliminability of the subjective factor in the development of science. Temperate rationalism of Newton-Smith was an attempt to solve this problem. J. Życiński developed his own version of temperate rationalism in which the subjective factor played much more substantial role. In the article I~am presenting his specific idea of the personal \textit{commitment} as a~necessary condition for rationalism and science. In the first section I~proceed to reconstruct Życiński's argument leading him to the conclusion of this epistemological necessity. Next in the section 2 I~present his idea of the epistemological uncertainty principle as a~consequence of the subjective \textit{commitment}. In sections 3 and 4 I~explore axiological and pragmatic aspects of the Życiński's solution. Finally, in the section 5 I~do compare his temperate rationalism with Newton-Smith's proposal in the context of the Polanyi's idea of personal knowledge showing differences in their respective approach to the role of the subjective factor in science and rationalism.}
	{subjective commitment, underdetermination principle, presuppositions, justification circle, principle of uncertainty, epistemic axiology, personal knowledge, temperate rationalism, pragmatic rationality, epistemology, doxalogy, voluntarism, Joseph Życiński.}

@@

Życiński wskazuje jednak
%\label{ref:RND5Ve53wlyzV}(Życiński, 1985, s.~133.161-164)
\parencite[][s.~133.161-164]{zycinski_teizm_1985} %
 na pewną brzemienną w~skutki metodologiczną różnicę pomiędzy empirycznym niedookreśleniem teorii w~naukach przyrodniczych a~analogicznym niedookreśleniem teorii w~metanauce. O~ile w~naukach przyrodniczych możliwe jest, pomimo teoretycznego obciążenia faktów, uzgadnianie stanowisk i~swoiste dążenie do unifikacji, o~tyle w~metanauce coś takiego wydaje się całkowicie niemożliwe. Wiara w~możliwość unifikacji metateoretycznej i~w możliwość powszechnej \textit{zgody filozofów} jest co najwyżej wyrazem nierealnego \textit{wishful thinking} i~pozostałością myślenia magicznego w~filozofii\footnote{Jako przykład pozostałości myślenia magicznego Życiński (1985, s.~133) podaje idee J. Monoda zawarte w~jego \textit{Le hasard et la nécessité} \parencite*{monod_hasard_1970}.}. Stoi ona w~sprzeczności z~ideą realistycznej koncepcji nauki respektującej dane historyczne:

@@

\textit{Zasadnicza niemożność} pozbycia się elementu subiektywnego z~nauki przekreśla tradycyjne wyobrażenia o~nauce jako dziele czystego rozumu. Wybór pomiędzy różnymi tradycjami badawczymi i~ich przedzałożeniami musi zostać uzupełniony o~akt podmiotowego \textit{związania} lub \textit{commitment}
%\label{ref:RND27y96wW0K5}(Życiński, 1985, s.~160.164nn.182, 1996, s.~186n.191, 2015, s.~253n.258)
\parencites[][s.~160]{zycinski_teizm_1985}[][s.~186n.191]{zycinski_elementy_1996}[][s.~253n.258]{zycinski_elementy_2015}%
\footnote{Życiński używa też określania ‘osobowe \textit{commitment}' pochodzącego zapewne od Polanyia 
%\label{ref:RNDCd2EdbIWIc}(1962),
\parencite*[][]{polanyi_personal_1962}, %
 zob. np. 
%\label{ref:RNDX0FNqDQW8z}(Życiński, 1985, s.~79.125.165).
\parencite[][s.~79.125.165]{zycinski_teizm_1985}. %
 Podobnie w
%\label{ref:RNDmC7GrxFJcP}(Życiński, 1988b, s.~137.144)
\parencite[][s.~137.144]{zycinski_structure_1988} %
 stosuje wyrażenia ‘\textit{subjective commitment}' i~‘\textit{personal commitment}', jak też zamiennie używa określenia ‘\textit{attachment}' (tamże, s.~143). W~pracach polskich Życiński stosuje termin angielski ‘\textit{commitment}' zamiennie z~polskimi odpowiednikami. Różne polskie znaczenia tego terminu Życiński przedstawia w
%\label{ref:RNDViGm1dQpm9}(Życiński, 1996, s.~191, 2015, s.~259n)
\parencites[][s.~191]{zycinski_elementy_1996}[][s.~259n]{zycinski_elementy_2015} %
 wskazując przy tym na T. Kuhna, W.V.O. Quine'a i~M. Polanyia jako tych, których prace przyczyniły się w~sposób szczególny do rozwoju badań nad rolą związania z~paradygmatem. Ponieważ angielskie \textit{commitment} stało się terminem technicznym, znacznie bardziej jednoznacznym od polskich odpowiedników -- np. \textit{związanie, zaangażowanie, przylgnięcie, zdeklarowanie} -- dlatego w~moim tekście pozostaję zasadniczo przy terminie ‘\textit{commitment}'.}:

@@

Oczywiście wszystkie argumenty, jakie przedstawia Życiński za realizmem, prawdą i~uprawdopodobnieniem mają charakter uzasadniania jedynie częściowego i~ostatecznie kolistego. Nie może to być jednak zarzutem, gdyż kolistość w~uzasadnianiu przedzałożeń jest istotnym elementem tradycji badawczej, z~którą wiąże się Życiński. W~sytuacji, gdy do wyboru mamy dwie zasadniczo równoważne tradycje racjonalności, Życiński opowiada się za tradycją, w~której racjonalność wiąże się z~prawdą jako celem poznania i~z umiarkowanym realizmem jako nową koncepcją tej prawdy. Wraz z~Newtonem-Smithem
%\label{ref:RNDF4NGv3QNVZ}(Newton-Smith, 1981, s.~273),
\parencite[][s.~273]{newton-smith_rationality_1981}, %
 a~zatem przeciw Laudanowi, z~którego koncepcją Newton-Smith polemizuje, konkluduje mówiąc, że ,,realizm jest prawdą a~umiarkowany racjonalizm -- drogą do prawdy'' 
%\label{ref:RNDTqCW4MDzzM}(Życiński, 1985, s.~205, 1988b, s.~135, 2013, s.~239)
\parencites[][s.~205]{zycinski_teizm_1985}[][s.~135]{zycinski_structure_1988}[][s.~239]{zycinski_struktura_2013_liana}%
\footnote{Podobne podejście do Laudanowskiej koncepcji celów nauki Laudana zaproponował wcześniej A. Grobler 
%\label{ref:RNDYps8mmrkOh}(1993, s.~23.35-42).
\parencite*[][s.~23.35-42]{grobler_prawda_1993}. %
 O~ile jednak Życiński omawia koncepcję Laudana w~kontekście \textit{verisimilitude}, o~tyle Grobler w~szerszym kontekście aksjologicznej luki.}.

>>>>

\begin{artplenv}{Zbigniew Liana}
	{Józefa Życińskiego koncepcja racjonalizmu umiarkowanego: epistemologiczna i~doksalogiczna funkcja podmiotowego \textit{commitment}}
	{Józefa Życińskiego koncepcja racjonalizmu umiarkowanego\ldots}
	{Józefa Życińskiego koncepcja racjonalizmu umiarkowanego: epistemologiczna i~doksalogiczna funkcja podmiotowego \tocname{commitment}}
	{Uniwersytet Papieski Jana Pawła II w~Krakowie\\	
	Centrum Kopernika Badań Interdyscyplinarnych\label{liana-anfang}}
	{Joseph Życiński's understanding of temperate rationalism: epistemological and doxalogical role of subjective commitment}
	{One of the main problems of modern rationalistic theories of science is the non-eliminability of the subjective factor in the development of science. Temperate rationalism of Newton-Smith was an attempt to solve this problem. J. Życiński developed his own version of temperate rationalism in which the subjective factor played much more substantial role. In the article I~am presenting his specific idea of the personal \textit{commitment} as a~necessary condition for rationalism and science. In the first section I~proceed to reconstruct Życiński's argument leading him to the conclusion of this epistemological necessity. Next in the section 2 I~present his idea of the epistemological uncertainty principle as a~consequence of the subjective \textit{commitment}. In sections 3 and 4 I~explore axiological and pragmatic aspects of the Życiński's solution. Finally, in the section 5 I~do compare his temperate rationalism with Newton-Smith's proposal in the context of the Polanyi's idea of personal knowledge showing differences in their respective approach to the role of the subjective factor in science and rationalism.}
	{subjective commitment, underdetermination principle, presuppositions, justification circle, principle of uncertainty, epistemic axiology, personal knowledge, temperate rationalism, pragmatic rationality, epistemology, doxalogy, voluntarism, Joseph Życiński.}

@@

Życiński wskazuje jednak
%\label{ref:RND5Ve53wlyzV}(Życiński, 1985, s.~133.161-164)
\parencite[][s.~133.161--164]{zycinski_teizm_1985} %
 na pewną brzemienną w~skutki metodologiczną różnicę pomiędzy empirycznym niedookreśleniem teorii w~naukach przyrodniczych a~analogicznym niedookreśleniem teorii w~metanauce. O~ile w~naukach przyrodniczych możliwe jest, pomimo teoretycznego obciążenia faktów, uzgadnianie stanowisk i~swoiste dążenie do unifikacji, o~tyle w~metanauce coś takiego wydaje się całkowicie niemożliwe. Wiara w~możliwość unifikacji metateoretycznej i~w możliwość powszechnej \textit{zgody filozofów} jest co najwyżej wyrazem nierealnego \textit{wishful thinking} i~pozostałością myślenia magicznego w~filozofii\footnote{Jako przykład pozostałości myślenia magicznego Życiński (1985, s.~133) podaje idee J. Monoda zawarte w~jego \textit{Le hasard et la nécessité} \parencite*{monod_hasard_1970}.}. Stoi ona w~sprzeczności z~ideą realistycznej koncepcji nauki respektującej dane historyczne:

@@

\textit{Zasadnicza niemożność} pozbycia się elementu subiektywnego z~nauki przekreśla tradycyjne wyobrażenia o~nauce jako dziele czystego rozumu. Wybór pomiędzy różnymi tradycjami badawczymi i~ich przedzałożeniami musi zostać uzupełniony o~akt podmiotowego \textit{związania} lub \textit{commitment}
%\label{ref:RND27y96wW0K5}(Życiński, 1985, s.~160.164nn.182, 1996, s.~186n.191, 2015, s.~253n.258)
\parencites[][s.~160]{zycinski_teizm_1985}[][s.~186n.191]{zycinski_elementy_1996}[][s.~253n.258]{zycinski_elementy_2015}%
\footnote{Życiński używa też określania ‘osobowe \textit{commitment}' pochodzącego zapewne od Polanyia 
%\label{ref:RNDCd2EdbIWIc}(1962),
\parencite*[][]{polanyi_personal_1962}, %
 zob. np. 
%\label{ref:RNDX0FNqDQW8z}(Życiński, 1985, s.~79.125.165).
\parencite[][s.~79.125.165]{zycinski_teizm_1985}. %
 Podobnie w
%\label{ref:RNDmC7GrxFJcP}(Życiński, 1988b, s.~137.144)
\parencite[][s.~137.144]{zycinski_structure_1988} %
 stosuje wyrażenia ‘\textit{subjective commitment}' i~‘\textit{personal commitment}', jak też zamiennie używa określenia ‘\textit{attachment}' (tamże, s.~143). W~pracach polskich Życiński stosuje termin angielski ‘\textit{commitment}' zamiennie z~polskimi odpowiednikami. Różne polskie znaczenia tego terminu Życiński przedstawia w
%\label{ref:RNDViGm1dQpm9}(Życiński, 1996, s.~191, 2015, s.~259n)
\parencites[][s.~191]{zycinski_elementy_1996}[][s.~259n]{zycinski_elementy_2015} %
 wskazując przy tym na T. Kuhna, W.V.O. Quine'a i~M.~Polanyia jako tych, których prace przyczyniły się w~sposób szczególny do rozwoju badań nad rolą związania z~paradygmatem. Ponieważ angielskie \textit{commitment} stało się terminem technicznym, znacznie bardziej jednoznacznym od polskich odpowiedników -- np. \textit{związanie, zaangażowanie, przylgnięcie, zdeklarowanie} -- dlatego w~moim tekście pozostaję zasadniczo przy terminie ‘\textit{commitment}'.}:

@@

Oczywiście wszystkie argumenty, jakie przedstawia Życiński za realizmem, prawdą i~uprawdopodobnieniem mają charakter uzasadniania jedynie częściowego i~ostatecznie kolistego. Nie może to być jednak zarzutem, gdyż kolistość w~uzasadnianiu przedzałożeń jest istotnym elementem tradycji badawczej, z~którą wiąże się Życiński. W~sytuacji, gdy do wyboru mamy dwie zasadniczo równoważne tradycje racjonalności, Życiński opowiada się za tradycją, w~której racjonalność wiąże się z~prawdą jako celem poznania i~z umiarkowanym realizmem jako nową koncepcją tej prawdy. Wraz z~Newtonem-Smithem
%\label{ref:RNDF4NGv3QNVZ}(Newton-Smith, 1981, s.~273),
\parencite[][s.~273]{newton-smith_rationality_1981}, %
 a~zatem przeciw Laudanowi, z~którego koncepcją Newton-Smith polemizuje, konkluduje mówiąc, że ,,realizm jest prawdą a~umiarkowany racjonalizm -- drogą do prawdy'' 
%\label{ref:RNDTqCW4MDzzM}(Życiński, 1985, s.~205, 1988b, s.~135, 2013, s.~239)
\parencites[][s.~205]{zycinski_teizm_1985}[][s.~135]{zycinski_structure_1988}[][s.~239]{zycinski_struktura_2013_liana}%
\footnote{Podobne podejście do Laudanowskiej koncepcji celów nauki Laudana zaproponował wcześniej A. Grobler 
%\label{ref:RNDYps8mmrkOh}(1993, s.~23.35-42).
\parencite*[][s.~23.35--42]{grobler_prawda_1993}. %
 O~ile jednak Życiński omawia koncepcję Laudana w~kontekście \textit{verisimilitude}, o~tyle Grobler w~szerszym kontekście aksjologicznej luki.}.
